---
title: "Fit width class data using SSN"
author: "Helen"
date: today
format: gfm
editor: visual
---


Step 1: 
Load the data

```{r load}
library(sf)
library(sfnetworks)
library(SSN2)
library(ggplot2)
library(data.table)

nodes <- read_sf("data/AreaC_Nodes/")
reaches <- read_sf("data/reaches/")
nodes$CLASS <-  as.numeric(factor(nodes$SurvType, 
                        levels = c("EPH", 
                                   "INT", 
                                   "TRANS", 
                                   "SMPRM", 
                                   "LGPRM")))
nodes$CLASS <- nodes$CLASS - 1

reaches$CLASS <- ifelse(reaches$CLASS %in% c(-9999, 0), NA, reaches$CLASS) - 1

# match up nodes to the nearest reach id
st_crs(nodes) <- st_crs(reaches)
nodes_matched <- st_join(nodes, reaches[c("REACH_ID")], all.x = TRUE, all.y = FALSE, join = st_nearest_feature)
names(reaches)
```

# fit a linear model. 

Here I fit both a binomial model and an ordinal model to compare outcomes. 

Note that binomial regression is generally used for a somewhat different 
scenario. It assumes that your response is a binomial random variable, which 
models the number of successes out of n trials where the probability of success 
for each trial is p. Binomial regression assumes that p varies with covariates, 
such that 

```{r}
data <- as.data.table(reaches[!is.na(reaches$CLASS), ])

# Start with ordinal regression 
m_ord <- MASS::polr(
  as.factor(CLASS) ~ log10(AREA_SQKM) + GRADIENT, 
  data = data, 
  Hess = TRUE)
summary(m_ord)
coef(m_ord)
preds_ord_y <- coef(m_ord)[1] * log10(data$AREA_SQKM) + 
  coef(m_ord)[2] * data$GRADIENT
preds_ord <- predict(m_ord)
hist(as.numeric(preds_ord))


# binomial regression? 
# at least it will return the correct data type, I guess

response <- matrix(c(data$CLASS, 3-data$CLASS), ncol = 2)

m_bin <- glm(response ~ log10(AREA_SQKM) + GRADIENT, family = binomial(link = "logit"), data = data)

summary(m_bin)
preds <- predict(m_bin, type = "response")

f_bin <- sapply(preds, \(p) dbinom(0:3, size = 3, prob = p)) |> t()
preds_bin <- sapply(preds, \(p) which.max(dbinom(0:3, size = 3, prob = p)))
hist(preds_bin)
hist(response[, 1])


```

### Compare model predictions. 
They are similar. For some reason ordinal regression is never predicting class 2. 
Some model fit stats are shown below, and a confusion matrix where the "reference" 
is the ordinal regression predictions and "prediction" is binomial predictions. 

The binomial model tends to predict lower than ordinal regression and never 
predicts higher than ordinal regression. 

```{r validate}
library(caret)
validation <- data.table(obs = data$CLASS, pred_bin = preds_bin-1, pred_ord = preds_ord, 
                         pred_bin_p = preds, 
                         pred_ord_y = preds_ord_y)
mx_ord <- confusionMatrix(as.factor(validation$obs), as.factor(validation$pred_ord))
mx_bin <- confusionMatrix(as.factor(validation$obs), as.factor(validation$pred_bin))

knitr::kable(data.table(
  model = c("Binomial", "Ordinal logistic"), 
  RMSE = c(sqrt(mean((validation$obs - validation$pred_bin)^2)), 
           sqrt(mean((validation$obs - as.numeric(validation$pred_ord))^2))), 
  accuracy = c(
    mx_ord$overall["Accuracy"],
    mx_bin$overall["Accuracy"])
))

boxplot(pred_bin_p ~ obs, data = validation, main = "Fitted binomial probabilites")
boxplot(pred_ord_y ~ obs, data = validation, main = "Fitted Y from ordinal regression")


confusionMatrix(as.factor(validation$pred_bin), as.factor(validation$pred_ord))

```

# Fit a spatial model

To do this, I first need to create a data structure which SSN can use. 
It will likely be faster to do this when exporting the synthetic network, but 
that functionality does not exist yet so I will just do it here. 

```{r}

```

```{r}
library(Matrix)
# Create adjacency matrix
# first need to index nodes from 1:n
n <- as.data.table(nodes)
setkey(n, NodeNum)
n$nid <- 1:nrow(n)
# connect to downstream node using new ids
# This uses the data.table key to find the row where NodeNum == ToNode, then get 
# the new nid, and assign that to a new to_nid column
n$to_nid <- n[.(ToNode), nid]
setkey(n, nid)
N <- nrow(n)

# edges
# This uses data.table's fast sort with keys to 
e <- n[, .(from = nid, to = to_nid)]
e <- e[!is.na(to)]


A <- sparseMatrix(i=e$from, 
                  j=e$to, 
                  x=1, 
                  dims=c(N,N) )

# start with only sources, outlets, and confluences
sources <- which(colSums(A) == 0)
outlets <- which(rowSums(A) == 0)
confluences <- which(colSums(A) == 2)


# create new data tables which will hold info for reduced network

# Start with node ids for only sources and confluences. Each of these 
# will correspond to one edge. There may be some confluences which are 
# also outlets. DO NOT include these because they won't have a downstream 
# edge associated with them. 

c <- confluences[!confluences %in% outlets]
new_n <- data.table(old_id = c(sources, c))
setorder(new_n, old_id)
new_n$new_id <- 1:nrow(new_n)
new_e <- data.table(from = new_n$new_id)
new_n <- rbind(new_n, data.table(old_id = outlets, new_id = 1:length(outlets) + nrow(new_n)))


# For each source or confluence node, get all nodes between it and the next 
# downstream confluence or outlet

# This takes a while for all 28k or so points so let's do it in parallel. 
library(parallel)
n_cores <- detectCores()
cl <- makeCluster(n_cores)
clusterEvalQ(cl, {
  library(data.table)
})

get_downstream <- function(id, i = 1, n_list = id) {
  to = n[id, to_nid]
  n_list <- c(n_list, to)
  i = i+1 
  # id <- to
  # to %in% c(confluences, outlets)
  if (to %in% c(confluences, outlets)) {
    return(list(to = new_n[old_id == to, new_id], i = i, n_list = n_list))
  } else {
    get_downstream(to, i, n_list)
  }
}


clusterExport(cl, c("n", "confluences", "outlets", "new_n", "get_downstream"))

sets <- parLapply(cl, new_n[new_e$from, old_id], get_downstream)


# Now use the list of nodes associated with each edge to get geometry for the 
# edge. 
n_sf <- st_as_sf(n)
clusterExport(cl, "n_sf")
clusterEvalQ(cl, {
  library(sf)
})

geometryList <- parLapply(cl, sets, \(s) {st_linestring(st_coordinates(n_sf)[s$n_list, ])}) 

new_e$geometry <- st_as_sfc(geometryList)
# record contributing area at most downstream point to use for SSN weighting
area_weights <- parLapply(cl, sets, \(s) {
  n[nid == s$n_list[length(s$n_list) - 1], AREA_SQKM]
})
new_e$area_weight <- unlist(area_weights)

new_e$to <- sapply(sets, \(s) s$to)
new_e$rid <- 1:nrow(new_e)
new_e <- st_as_sf(new_e, crs = st_crs(reaches))

# get nids so we can get info from nodes table later
 
setkey(new_n, new_id)
new_e$from_nodenum <- n[.(new_n[.(new_e$from), old_id]), NodeNum]
new_e$to_nodenum <- n[.(new_n[.(new_e$to), old_id]), NodeNum]

# get nids of center node of edge for predictions
new_e$center_nodenum <- n[.(unlist(lapply(sets, \(s) s$n_list[round(s$i/2)]))), NodeNum]

stopCluster(cl)
```

```{r make_ssn}
library(SSN2)
library(SSNbler)

# predict at center(ish) node for each edge
pred_points <- st_as_sf(n[.(unlist(new_e$center_nid))])

# set observation points as center point from each reach
# https://gis.stackexchange.com/questions/254151/how-to-find-the-center-point-which-lies-on-the-linestring-geometry 
reach_obs <- reaches[!is.na(reaches$CLASS), ]

reach_obs$geometry <- st_line_interpolate(st_geometry(reach_obs), 0.5, normalized = TRUE)
reach_obs

# Need to convert from data.table to data.frame (weird bugs happen otherwise)
if (!dir.exists("ssn_files_keep/ssn.ssn")) {
  lsn.path <- file.path(tempdir(), "stream_width")
  
  edges <- lines_to_lsn(
    streams = st_as_sf(as.data.frame(new_e), crs = st_crs(new_e)),
    lsn_path = lsn.path,
    check_topology = TRUE,
    snap_tolerance = 1,
    topo_tolerance = 20,
    overwrite = TRUE
  )
  obs <- sites_to_lsn(
    sites = st_as_sf(reach_obs, crs = st_crs(reach_obs)),
    edges = edges,
    lsn_path = lsn.path,
    file_name = "obs",
    snap_tolerance = 1,
    save_local = TRUE,
    overwrite = TRUE
  )
  preds <- sites_to_lsn(
    sites = st_as_sf(as.data.frame(pred_points), crs = st_crs(pred_points)),
    edges = edges,
    save_local = TRUE,
    lsn_path = lsn.path,
    file_name = "pred",
    snap_tolerance = 1,
    overwrite = TRUE
  )
  
  edges <- updist_edges(
    edges = edges,
    save_local = TRUE,
    lsn_path = lsn.path,
    calc_length = TRUE
  )
  
  site.list <- updist_sites(
    sites = list(
      obs = obs,
      preds = preds
    ),
    edges = edges,
    length_col = "Length",
    save_local = TRUE,
    lsn_path = lsn.path
  )
  
  edges <- afv_edges(
    edges = edges,
    infl_col = "area_weight",
    segpi_col = "areaPI",
    afv_col = "afvArea",
    lsn_path = lsn.path
  )
  
  site.list <- afv_sites(
    sites = site.list,
    edges = edges,
    afv_col = "afvArea",
    save_local = TRUE,
    lsn_path = lsn.path
  )
  reaches_ssn <- ssn_assemble(
    edges = edges,
    lsn_path = lsn.path,
    obs_sites = site.list$obs,
    preds_list = site.list[c("preds")],
    ssn_path = "ssn_files_keep/ssn",
    import = TRUE,
    check = TRUE,
    afv_col = "afvArea",
    overwrite = TRUE
  )
} else {
  reaches_ssn <- ssn_import("ssn_files_keep/ssn.ssn", predpts = "preds", overwrite = TRUE)
}

```





