---
title: "Fit width class data using SSN"
author: "Helen"
date: today
output: github_document
editor: source
---

Step 1: Load the data

```{r load}
library(sf)
library(sfnetworks)
library(SSN2)
library(ggplot2)
library(data.table)

nodes <- read_sf("data/AreaC_Nodes/")
reaches <- read_sf("data/reaches/")
nodes$CLASS <-  as.numeric(factor(nodes$SurvType, 
                        levels = c("EPH", 
                                   "INT", 
                                   "TRANS", 
                                   "SMPRM", 
                                   "LGPRM")))
nodes$CLASS <- nodes$CLASS - 1

reaches$CLASS <- ifelse(reaches$CLASS %in% c(-9999, 0), NA, reaches$CLASS) - 1

# match up nodes to the nearest reach id
st_crs(nodes) <- st_crs(reaches)
nodes_matched <- st_join(nodes, reaches[c("REACH_ID")], all.x = TRUE, all.y = FALSE, join = st_nearest_feature)
names(reaches)
```

# fit a linear model.

Here I fit both a binomial model and an ordinal model to compare outcomes.

Note that binomial regression is generally used for a somewhat different scenario. It assumes that your response is a binomial random variable, which models the number of successes out of n trials where the probability of success for each trial is p. Binomial regression assumes that p varies with covariates, such that

```{r fit-nonspatial-models}
data <- as.data.table(reaches[!is.na(reaches$CLASS), ])

# Start with ordinal regression 
m_ord <- MASS::polr(
  as.factor(CLASS) ~ log10(AREA_SQKM) + GRADIENT, 
  data = data, 
  Hess = TRUE)
summary(m_ord)
coef(m_ord)
preds_ord_y <- coef(m_ord)[1] * log10(data$AREA_SQKM) + 
  coef(m_ord)[2] * data$GRADIENT
preds_ord <- predict(m_ord)
hist(as.numeric(preds_ord))


# binomial regression? 
# at least it will return the correct data type, I guess

response <- matrix(c(data$CLASS, 3-data$CLASS), ncol = 2)

m_bin <- glm(response ~ log10(AREA_SQKM) + GRADIENT, family = binomial(link = "logit"), data = data)

summary(m_bin)
preds <- predict(m_bin, type = "response")

f_bin <- sapply(preds, \(p) dbinom(0:3, size = 3, prob = p)) |> t()
preds_bin <- sapply(preds, \(p) which.max(dbinom(0:3, size = 3, prob = p)))
hist(preds_bin)
hist(response[, 1])


```

### Compare model predictions.

They are similar. For some reason ordinal regression is never predicting class 2. Some model fit stats are shown below, and a confusion matrix where the "reference" is the ordinal regression predictions and "prediction" is binomial predictions.

The binomial model tends to predict lower than ordinal regression and never predicts higher than ordinal regression.

```{r validate}
library(caret)
validation <- data.table(obs = data$CLASS, pred_bin = preds_bin-1, pred_ord = preds_ord, 
                         pred_bin_p = preds, 
                         pred_ord_y = preds_ord_y)
mx_ord <- confusionMatrix(as.factor(validation$obs), as.factor(validation$pred_ord))
mx_bin <- confusionMatrix(as.factor(validation$obs), as.factor(validation$pred_bin))

knitr::kable(data.table(
  model = c("Binomial", "Ordinal logistic"), 
  RMSE = c(sqrt(mean((validation$obs - validation$pred_bin)^2)), 
           sqrt(mean((validation$obs - as.numeric(validation$pred_ord))^2))), 
  accuracy = c(
    mx_ord$overall["Accuracy"],
    mx_bin$overall["Accuracy"])
))

boxplot(pred_bin_p ~ obs, data = validation, main = "Fitted binomial probabilites")
boxplot(pred_ord_y ~ obs, data = validation, main = "Fitted Y from ordinal regression")


confusionMatrix(as.factor(validation$pred_bin), as.factor(validation$pred_ord))

```

# Fit a spatial model

Load and explore the .ssn object. 
The torgegram shows the semivariance for three different types of distance:  
- flowcon: stream distance betweeen flow-connected points
- flowuncon: stream distance between all (flow-connected and -unconnected points)
- euclid: Euclidean distance

The scale of variance seems to be different between flow-connected points and 
others, so we will plot them separately. 

```{r load-ssn}
library(SSN2)
ssn <-  ssn_import(
  path = "data/ssn_files_keep/ssn.ssn",
  predpts = c("preds"), 
  overwrite = TRUE
)

tg_connected <- Torgegram(
  formula = CLASS ~ log10(AREA_SQKM) + GRADIENT,
  ssn.object = ssn,
  type = c("flowcon")
)
plot(tg_connected)
tg_unconnected <- Torgegram(
  formula = CLASS ~ log10(AREA_SQKM) + GRADIENT,
  ssn.object = ssn,
  type = c("flowuncon", "euclid")
)
plot(tg_unconnected)
```

There clearly is increasing variance between flow-connected points. I don't really see this pattern in flow-unconnected points but there is some increasing variance between euclidan distance. 
This indicates that a tail-up variance component would be appropriate, and potentially also a component 
for euclidean distance.

Unfortunately, I am getting an error when I attempt to fit a tail-up spatial effect. I'm not 
sure why, but it may be related to the large number of separated networks. Instead, let's fit a 
tail-down effect. 

```{r fit-ssn}
# this took about 30 min
ssn_mod_binom <- ssn_glm(
  matrix(c(CLASS, 3-CLASS), ncol = 2) ~ log10(AREA_SQKM) + GRADIENT, 
                   family = "binomial",
                   ssn,  
                   taildown_type = "exponential",
                   additive = "afvArea")

# ssn_mod_binom <- ssn_glm(CLASS ~ log10(AREA_SQKM) + GRADIENT, 
#                    family = "binomial",
#                    size = 3,
#                    ssn,  
#                    taildown_type = "exponential",
#                    additive = "afvArea")
# t3 <- Sys.time()
# ssn_mod_gamma <- ssn_glm(CLASS ~ log10(AREA_SQKM) + GRADIENT, 
#                          family = "gamma",
#                          ssn,  
#                          taildown_type = "exponential",
#                          additive = "afvArea")


```

```{r ssn-summary}
summary(ssn_mod_binom)
```

Parameter estimates are pretty different between the spatial and non-spatial binomial models. 
Let's look at fitted values. 

```{r validate-ssn}
# for some reason they get transformed to be between 0 and 3 
# when you say type = "response"
p_fitted_ssn <- plogis(fitted(ssn_mod_binom, type = "link"))
f_ssn <- sapply(p_fitted_ssn, \(p) dbinom(0:3, size = 3, prob = p)) |> t()
preds_ssn <- sapply(p_fitted_ssn, \(p) which.max(dbinom(0:3, size = 3, prob = p)))

hist(preds_ssn)
boxplot(p_fitted_ssn ~ ssn$obs$CLASS)

ssn_mx <- confusionMatrix(as.factor(ssn$obs$CLASS), as.factor(preds_ssn -1))
sqrt(mean((ssn$obs$CLASS - (preds_ssn -1))^2))

knitr::kable(data.table(
  model = c("Binomial", "Ordinal logistic", "SSN (binomial)"), 
  RMSE = c(sqrt(mean((validation$obs - validation$pred_bin)^2)), 
           sqrt(mean((validation$obs - as.numeric(validation$pred_ord))^2)), 
           sqrt(mean((ssn$obs$CLASS - (preds_ssn -1))^2))), 
  accuracy = c(
    mx_ord$overall["Accuracy"],
    mx_bin$overall["Accuracy"], 
    ssn_mx$overall["Accuracy"])
))
```

This is a pretty significant improvement in fitted values. However, it is not perfect, despite 
the spatial fitting we are doing. There are likely a couple of reasons. The first is that 
the spatial smoothing that is included in the model fit is a smoother, it doesn't snap values 
exactly to observations. The second is the type of model we are using. We are using binomial 
regression, which is different from ordinal regression but can be fit to data which looks similar 
(integer data within a set range). Rather than predict an ordinal class, binomial regression predicts 
a probability. We can then translate that probability into a predicted number, but that's not really 
what it's designed for. 

Another implication of using binomial rather than ordinal regression is that it will be difficult to 
get sensible estimates of uncertainty from the model output. We can use uncertainty estimates for regression parameters but there is no variance or dispersion parameter in binomial regression which 
can be fit to describe variation around the predicted mean. Quasi-binomial (which is built in to the 
glm function) or ordinal regression does fit a dispersion parameter. Without modifying the SSN2 
codebase to allow fitting ordinal or quasibinomial models, we might be able to back out a bootstrapped 
estimate of prediction uncertianty. 
