---
title: "Ordinal logistic regression"
output: github_document
---

```{r setup}
suppressPackageStartupMessages({
  library(sf)
  library(terra)
  library(tmap)
  library(data.table)
  library(ggplot2)
  library(PResiduals)
  library(sure)
})

# elev <- rast("streams_clipped/elev_areactest.tif")
nodes <- read_sf("streams_clipped/nodes_testc.shp")
cut_blocks <- read_sf("streams_clipped/cut_blocks.shp")
streams <- read_sf("streams_clipped/streams_clipped.shp")
```

```{r clean_data}

# Make sure everything is in the same CRS
# set everything to utm 11N
cut_blocks <- st_transform(cut_blocks, "EPSG:26911")
#elev <- project(elev, "EPSG:26911")
st_crs(nodes) <- "EPSG:26911"

nodes[nodes$SurvReach == -9999, ]$SurvReach <- NA

survey_nodes <- nodes[!is.na(nodes$SurvType) & 
                        nodes$SurvType != "UNKN", ]


# # choose one cut block to work with
# block_id <- "S250922"
# block <- cut_blocks[cut_blocks$BLOCK_ID == block_id, ]
# 
# # trim nodes and streams 
# nodes <- nodes[block, ]
# streams_block <- streams[block, ]
```

In this notebook, I fit an ordinal model to a subset of the data and evaluate some model fit diagnostics. 

Here is what the data looks like: 

```{r map, eval = F}

# tm_shape(cut_blocks) +
#   tm_lines() +
#   tm_shape(nodes) + 
#   tm_dots(size = 0.05) +
#   tm_shape(streams) + 
#   tm_lines(col = "RIPARIAN_C", lwd = 2)

```

This is the distribution of all nodes in the network vs the nodes which we have data for. 

Survey nodes skew right. I wonder if this may be because nodes with low contributing area often did not show up as channels (ie width = 0). 


```{r density-plots}

data <- nodes
st_geometry(data) <- NULL
setDT(data)

data <- data[, .(
  NodeNum, 
  stream_class = factor(SurvType, 
                        levels = c("EPH", 
                                   "INT", 
                                   "TRANS", 
                                   "SMPRM", 
                                   "LGPRM")), 
  area = AREA_SQKM, 
  area_log10 = log10(AREA_SQKM), 
  elev = ELEV_M, 
  grad20 = GRAD20, 
  mean_grad = MeanGrad
)]

data_md <- data[!is.na(stream_class)]


ggplot() +
  geom_density(data = data, aes(x = area, fill = "All"), alpha = 0.4) +
  geom_density(data = data_md, aes(x = area, fill = "Survey"), alpha = 0.4) +
  scale_x_continuous(trans = "log10") 

ggplot(data_md) + 
  geom_boxplot(aes(x = stream_class, y = area)) + 
  scale_y_log10()

```

```{r fit-model1}

# rms::lrm wasn't working for some reason
# but this presumably does the same thing
# (ordered logistic regression)
m1 <- MASS::polr(
  stream_class ~ area_log10, 
  data = data_md, 
  Hess = TRUE)
summary(m1)
f <- fitted(m1)
head(f)
apply(f, 2, max)
apply(f, 2, sum)

```

# Residual analysis

First, we get residuals and plot them using SBS residuals in the `PResiduals` package. 

```{r sbs-resid}
# get resid using SBS
resid_sbs <- presid(m1)

par(mfcol = c(1, 2))
plot(x = data_md$area_log10, y = resid_sbs, pch = 2)
qqplot(runif(length(resid_sbs)), resid_sbs)
```

Next, we use surrogate residuals from the `sure` package. 

```{r sure-resid-1}
resid_sure <- sure::resids(m1)

par(mfcol = c(1, 2))
plot(x = data_md$area_log10, y = resid_sure, pch = 2)
qqnorm(resid_sure)
```

They look a bit funky, but it might be because our data is not evenly distributed between classes. Let's test for spatial autocorrelation. 

```{r plot-sure-resid-1}
# plot residuals
data_md$resid <- resid_sure
merged <- merge(data_md, nodes, by = "NodeNum", all.x = TRUE, all.y = FALSE) |> 
  st_as_sf()

tm_shape(merged) + 
  tm_dots(fill = "resid", 
          fill.scale = tm_scale_continuous())
```

```{r spatial-autocorrelation-1}
# Spatial autocorrelation in resiuals
library(gstat)
library(spdep)

# semivariogram
v <- variogram(resid ~ 1, merged)
plot(v)

# Moran's I
# Set k nearest neighbors
nb <- knn2nb(knearneigh(merged, 1))
lw <- nb2listw(nb)
lag <- lag.listw(lw, merged$resid)
plot(merged$resid, lag)
moran.test(merged$resid, lw)
```

Based on the Moran's I test, we have significant spatial autocorrelation in the 
surrogate residuals. 
One issue here is that the observations are on a reach level while our 
predictions are at a node level. Let's try again, aggregating data by reach. 

To do this, I will create a sfnetwork object from the sf nodes object. 
Then, I will aggregate nodes into reaches by defining a reach as a length 
of stream between two confluences, and aggregating covariates. 
I will only include reaches as predictors where all nodes within that reach 
were assigned the same stream class. 

This is an imperfect way of doing this, but hopefully should be enough 
to get a rough idea of whether we are still seeing spatial autocorrelation 
or whether it was just due primarily to different aggregation levels. 

```{r aggregate-reaches}
library(igraph)
library(tidygraph)
library(sfnetworks)
setDT(nodes)
#nodes <- st_as_sf(nodes)
edges <- nodes[, .(from = NodeNum, 
                    to = ToNode, 
                    surv_type = SurvType,
                    channel_id = CHANNEL_ID, 
                    area = AREA_SQKM, 
                    elev = ELEV_M, 
                    grad = GRAD20, 
                    width_m = WIDTH_M, 
                    depth_m = DEPTH_M, 
                    incise = INCISE10
                    )]
edges[to == -9999, to := NA]
edges <- edges[!is.na(to)]

channel_network <- sfnetwork(nodes, edges)
# add geometry to edges (draw a straight line between nodes)
channel_network = convert(channel_network, to_spatial_explicit)
# aggregate by reach (remove intermediate nodes and concatenate edges)
str_unique <- function(x) paste(unique(na.omit(x), na.rm= TRUE), collapse = ",")
channel_reaches <- convert(channel_network, to_spatial_smooth, 
                           summarise_attributes = list(
                             channel_id = str_unique, 
                             surv_type = str_unique,
                             area = "max", 
                             elev = "mean", 
                             grad = "mean", 
                             width_m = "mean", 
                             depth_m = "mean", 
                             incise = "mean"))

edges_reaches <- st_as_sf(activate(channel_reaches, "edges"))
edges_reaches$.tidygraph_edge_index <- NULL
st_crs(edges_reaches) <- "EPSG:26911"
# write to shapefile
# st_write(edges_reaches, "edges_reaches/edges_reaches.shp")
```

```{r fit-model-2}

# make ordered factor stream class for ordinal regression
# remove NA
unique_class <- function(s) {
  st <- strsplit(s, ",")[[1]]
  st <- st[!st == "NA"]
  if (length(st) == 1) return(st) else return(NA)
}
edges_reaches$stream_class <- factor(
  sapply(edges_reaches$surv_type, unique_class), 
  levels = c("EPH", 
             "INT", 
             "TRANS", 
             "SMPRM", 
             "WET"))


data_m2 <- edges_reaches[!is.na(edges_reaches$stream_class), ]
data_m2$area_log10 <- log10(data_m2$area)

m2 <- MASS::polr(
  stream_class ~ area_log10, 
  data = data_m2, 
  Hess = TRUE)
summary(m2)
f <- fitted(m2)
head(f)
apply(f, 2, max)
apply(f, 2, sum)
```

```{r resid-2}
data_m2$resid <- sure::resids(m2)
par(mfcol = c(1, 2))
plot(x = data_m2$area_log10, y = data_m2$resid, pch = 2)
qqnorm(data_m2$resid)
```

```{r map-resid-2}
tm_shape(data_m2) + 
  tm_dots(fill = "resid", 
          fill.scale = tm_scale_continuous())
```

```{r spatial-autocorrelation-2}

# semivariogram 
# this isn't working with lines, only points
# use the node/point at the beginning of each reach
nodes_reaches <- st_as_sf(activate(channel_reaches, "nodes"))[edges_reaches$from, ]
data_nodes <- nodes_reaches[!is.na(edges_reaches$stream_class), ]
data_nodes$resid <- data_m2$resid
v <- variogram(resid ~ 1, data_nodes)
plot(v)

# Moran's I
# Set k nearest neighbors
nb <- knn2nb(knearneigh(data_nodes, 1))
lw <- nb2listw(nb)
lag <- lag.listw(lw, data_nodes$resid)
plot(data_nodes$resid, lag)
moran.test(data_nodes$resid, lw)
```

We are still seeing spatial autocorrelation. Next steps could include trying to
fit a model which includes more predictors and to add a spatial autocorrelation 
term. 
